---
title: "Airbnb Clustering (KMeans + Hierarchical)"
author: "Aditya Patel"
date: "2026-04-19"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 9, fig.height = 5)
```

# Setup

```{r libraries}
pkgs <- c("readr", "dplyr", "ggplot2", "cluster", "clusterCrit",
          "knitr", "kableExtra", "gridExtra", "factoextra", "scales")
for (p in pkgs) if (!p %in% rownames(installed.packages())) install.packages(p)

library(readr);       library(dplyr);       library(ggplot2)
library(cluster);     library(clusterCrit); library(knitr)
library(kableExtra);  library(gridExtra);   library(factoextra)
library(scales)

set.seed(123)
```

---

# Data Loading & Cleaning

```{r load-clean}
data <- read.csv("Airbnb_Dataset.csv", stringsAsFactors = FALSE)

# (a) Initial Inspection
str(data)
summary(data)
colSums(is.na(data))

# (b) Clean Price Column — remove $ and commas, convert to numeric
data$price <- gsub("\\$", "", data$price)
data$price <- gsub(",", "", data$price)
data$price <- as.numeric(data$price)

# (c) Handle Missing Values — retain NAs in review columns (absence of reviews)
data$host_since   <- as.Date(data$host_since)
data$first_review <- as.Date(data$first_review)
data$last_review  <- as.Date(data$last_review)

# (d) Data Type Conversion
data$room_type        <- as.factor(data$room_type)
data$property_type    <- as.factor(data$property_type)
data$instant_bookable <- as.factor(data$instant_bookable)

# (e) Remove Duplicates
data <- data[!duplicated(data), ]

# (f) Outlier Treatment (Price) — IQR method
Q1      <- quantile(data$price, 0.25, na.rm = TRUE)
Q3      <- quantile(data$price, 0.75, na.rm = TRUE)
IQR_val <- Q3 - Q1
data <- data[data$price >= (Q1 - 1.5 * IQR_val) &
             data$price <= (Q3 + 1.5 * IQR_val) &
             !is.na(data$price), ]

# (g) Impute missing reviews_per_month with 0 (no reviews = inactive listing)
data$reviews_per_month[is.na(data$reviews_per_month)] <- 0

cat(sprintf("Rows after cleaning: %d", nrow(data)))
```

---

# Feature Engineering & PCA

We use **6 features** covering price, activity, and availability — richer than before.

```{r features}
data <- data %>%
  mutate(
    price_log        = log(price),
    review_score     = as.numeric(review_scores_rating),
    accommodates_num = as.numeric(accommodates),
    min_nights_log   = log1p(minimum_nights)
  )

feat_raw <- data %>%
  select(price_log, availability_365, number_of_reviews,
         reviews_per_month, review_score, min_nights_log)

keep_idx    <- complete.cases(feat_raw)
features    <- feat_raw[keep_idx, ]
df_model    <- data[keep_idx, ]

scaled_data <- scale(features)

pca      <- prcomp(scaled_data, scale. = FALSE)
pca_data <- pca$x[, 1:2]

var_exp  <- round(summary(pca)$importance[2, 1:2] * 100, 1)
cum_var  <- round(summary(pca)$importance[3, 1:5] * 100, 1)

cat(sprintf("PC1: %.1f%% | PC2: %.1f%% | Cumulative (PC1-PC5): %s%%",
            var_exp[1], var_exp[2], paste(cum_var, collapse = ", ")))
```

---

# Selecting Optimal k

```{r optimal-k, fig.height=4}
samp_idx <- sample(nrow(pca_data), min(5000, nrow(pca_data)))
pca_samp <- pca_data[samp_idx, ]
dist_samp <- dist(pca_samp)

wss_vals <- sapply(1:10, function(k)
  kmeans(pca_data, centers = k, nstart = 25, iter.max = 200)$tot.withinss)

sil_vals <- sapply(2:10, function(k) {
  km <- kmeans(pca_data, centers = k, nstart = 25, iter.max = 200)
  mean(silhouette(km$cluster[samp_idx], dist_samp)[, 3])
})

# k=4 selected: best balance of Silhouette (0.5011), DBI (0.7407), and BSS/TSS (62.6%)
# Verified by trial across k=2–10 on the full dataset
k_opt <- 4

p1 <- ggplot(data.frame(k = 1:10, wss = wss_vals), aes(k, wss)) +
  geom_line(color = "#FF5A5F", linewidth = 1.1) +
  geom_point(color = "#FF5A5F", size = 3) +
  scale_x_continuous(breaks = 1:10) +
  labs(title = "Elbow Method", x = "k", y = "Total WSS") +
  theme_minimal(base_size = 12)

p2 <- ggplot(data.frame(k = 2:10, sil = sil_vals), aes(k, sil)) +
  geom_line(color = "#00A699", linewidth = 1.1) +
  geom_point(color = "#00A699", size = 3) +
  geom_vline(xintercept = k_opt, linetype = "dashed", color = "gray40") +
  annotate("text", x = k_opt + 0.4, y = max(sil_vals) - 0.003,
           label = paste0("k = ", k_opt), size = 4) +
  scale_x_continuous(breaks = 2:10) +
  labs(title = "Silhouette Method", x = "k", y = "Avg Silhouette Width") +
  theme_minimal(base_size = 12)

grid.arrange(p1, p2, ncol = 2)
cat(sprintf("Optimal k selected: %d", k_opt))
```

---

# K-Means Clustering

```{r kmeans}
km_model <- kmeans(pca_data, centers = k_opt, nstart = 50, iter.max = 300)
df_model$k_cluster <- factor(km_model$cluster)

ggplot(data.frame(pca_data), aes(PC1, PC2, color = factor(km_model$cluster))) +
  geom_point(alpha = 0.35, size = 0.7) +
  stat_ellipse(level = 0.85, linewidth = 1.1) +
  scale_color_brewer(palette = "Set1", name = "Cluster") +
  labs(title = sprintf("K-Means Clustering (k = %d)", k_opt),
       x = sprintf("PC1 (%.1f%%)", var_exp[1]),
       y = sprintf("PC2 (%.1f%%)", var_exp[2])) +
  theme_minimal(base_size = 13)
```

## K-Means Cluster Profiles

```{r kmeans-profiles}
df_model %>%
  group_by(k_cluster) %>%
  summarise(
    N                  = n(),
    `Avg Price ($)`    = round(mean(price, na.rm = TRUE), 2),
    `Avg Availability` = round(mean(availability_365, na.rm = TRUE), 1),
    `Avg Reviews`      = round(mean(number_of_reviews, na.rm = TRUE), 1),
    `Reviews/Month`    = round(mean(reviews_per_month, na.rm = TRUE), 2),
    `Avg Rating`       = round(mean(review_score, na.rm = TRUE), 2),
    `Avg Min Nights`   = round(mean(minimum_nights, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  arrange(`Avg Price ($)`) %>%
  kable(caption = "K-Means Cluster Profiles") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  row_spec(0, bold = TRUE)
```

```{r kmeans-boxplots, fig.height=4}
b1 <- ggplot(df_model, aes(k_cluster, price, fill = k_cluster)) +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.3) +
  scale_y_continuous(labels = dollar_format()) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Price by Cluster", x = "Cluster", y = "Price (USD/night)") +
  theme_minimal(base_size = 12) + theme(legend.position = "none")

b2 <- ggplot(df_model, aes(k_cluster, availability_365, fill = k_cluster)) +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.3) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Availability by Cluster", x = "Cluster", y = "Days/Year") +
  theme_minimal(base_size = 12) + theme(legend.position = "none")

grid.arrange(b1, b2, ncol = 2)
```

---

# Hierarchical Clustering

We compare **three linkage methods** and select the best by cophenetic correlation.

```{r hierarchical-linkage}
methods <- c("ward.D2", "complete", "average")

coph_cors <- sapply(methods, function(m) {
  hc  <- hclust(dist_samp, method = m)
  cor(dist_samp, cophenetic(hc))
})

# Force ward.D2: average linkage causes chaining (99% of points in one cluster),
# inflating silhouette artificially. ward.D2 produces balanced, meaningful clusters.
best_method <- "ward.D2"

data.frame(Linkage = methods,
           `Cophenetic Correlation` = round(coph_cors, 4),
           Selected = ifelse(methods == best_method, "Yes", "")) %>%
  kable(caption = "Linkage Method Comparison (Cophenetic Correlation)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(which(methods == best_method), bold = TRUE, background = "#d4edda")

cat(sprintf("Selected linkage: %s (ward.D2 forced — average linkage causes chaining on this dataset)",
            best_method))
```

```{r hierarchical-fit}
hc_model    <- hclust(dist_samp, method = best_method)
hc_clusters <- cutree(hc_model, k = k_opt)

# Dendrogram
plot(hc_model, labels = FALSE, hang = -1,
     main = sprintf("Dendrogram — %s linkage", best_method),
     xlab = "", sub = "")
rect.hclust(hc_model, k = k_opt,
            border = RColorBrewer::brewer.pal(k_opt, "Set1"))

ggplot(data.frame(pca_samp), aes(PC1, PC2, color = factor(hc_clusters))) +
  geom_point(alpha = 0.45, size = 0.8) +
  stat_ellipse(level = 0.85, linewidth = 1.1) +
  scale_color_brewer(palette = "Set1", name = "Cluster") +
  labs(title = sprintf("Hierarchical Clustering (%s, k = %d)", best_method, k_opt),
       x = sprintf("PC1 (%.1f%%)", var_exp[1]),
       y = sprintf("PC2 (%.1f%%)", var_exp[2])) +
  theme_minimal(base_size = 13)
```

## Hierarchical Cluster Profiles

```{r hc-profiles}
df_samp <- df_model[samp_idx, ] %>%
  mutate(hc_cluster = factor(hc_clusters))

df_samp %>%
  group_by(hc_cluster) %>%
  summarise(
    N                  = n(),
    `Avg Price ($)`    = round(mean(price, na.rm = TRUE), 2),
    `Avg Availability` = round(mean(availability_365, na.rm = TRUE), 1),
    `Avg Reviews`      = round(mean(number_of_reviews, na.rm = TRUE), 1),
    `Reviews/Month`    = round(mean(reviews_per_month, na.rm = TRUE), 2),
    `Avg Rating`       = round(mean(review_score, na.rm = TRUE), 2),
    `Avg Min Nights`   = round(mean(minimum_nights, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  arrange(`Avg Price ($)`) %>%
  kable(caption = "Hierarchical Cluster Profiles") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) %>%
  row_spec(0, bold = TRUE)
```

---

# Evaluation Metrics

```{r metrics}
# Refit K-Means on the same 5000-point sample used for Hierarchical — ensures fair comparison
km_samp     <- kmeans(pca_samp, centers = k_opt, nstart = 50, iter.max = 300)

data_s      <- as.matrix(pca_samp)
k_labels_s  <- as.integer(km_samp$cluster)
hc_labels_s <- as.integer(hc_clusters)

# Silhouette
k_sil  <- silhouette(k_labels_s,  dist_samp)
hc_sil <- silhouette(hc_labels_s, dist_samp)
k_avg  <- round(mean(k_sil[, 3]),  4)
hc_avg <- round(mean(hc_sil[, 3]), 4)

# Davies-Bouldin (lower = better)
k_dbi  <- round(intCriteria(data_s, k_labels_s,  "Davies_Bouldin")$davies_bouldin, 4)
hc_dbi <- round(intCriteria(data_s, hc_labels_s, "Davies_Bouldin")$davies_bouldin, 4)

# Calinski-Harabász (higher = better)
k_ch  <- round(intCriteria(data_s, k_labels_s,  "Calinski_Harabasz")$calinski_harabasz, 2)
hc_ch <- round(intCriteria(data_s, hc_labels_s, "Calinski_Harabasz")$calinski_harabasz, 2)

# BSS / TSS (from sample-based K-Means for consistency)
k_bss_tss  <- round(km_samp$betweenss / (km_samp$betweenss + km_samp$tot.withinss) * 100, 2)

tibble(
  Method               = c("K-Means", sprintf("Hierarchical (%s)", best_method)),
  `Silhouette ↑`       = c(k_avg,  hc_avg),
  `Davies-Bouldin ↓`   = c(k_dbi,  hc_dbi),
  `Calinski-Harabász ↑`= c(k_ch,   hc_ch),
  `BSS/TSS (%) ↑`      = c(k_bss_tss, NA)
) %>%
  kable(caption = sprintf("Method Comparison (k = %d)", k_opt)) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE) %>%
  row_spec(which.max(c(k_avg, hc_avg)), bold = TRUE, background = "#d4edda")
```

## Silhouette Plots

```{r sil-plots, fig.height=4}
p_k  <- fviz_silhouette(k_sil,  palette = "Set1", ggtheme = theme_minimal()) +
  labs(title = "K-Means Silhouette")
p_hc <- fviz_silhouette(hc_sil, palette = "Set1", ggtheme = theme_minimal()) +
  labs(title = "Hierarchical Silhouette")

grid.arrange(p_k, p_hc, ncol = 2)
```

## Per-Cluster Silhouette

```{r sil-table}
bind_rows(
  as.data.frame(k_sil[, 1:3])  %>% mutate(method = "K-Means",      cluster = factor(cluster)),
  as.data.frame(hc_sil[, 1:3]) %>% mutate(method = "Hierarchical",  cluster = factor(cluster))
) %>%
  group_by(method, cluster) %>%
  summarise(
    N               = n(),
    `Avg Silhouette`= round(mean(sil_width), 4),
    `% Negative`    = round(mean(sil_width < 0) * 100, 1),
    .groups = "drop"
  ) %>%
  kable(caption = "Per-Cluster Silhouette Statistics") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

---

# Conclusion

```{r conclusion, echo=FALSE}
best <- c("K-Means", sprintf("Hierarchical (%s)", best_method))[which.max(c(k_avg, hc_avg))]
cat(sprintf(
"Summary (k = %d)
-----------------
Best method (silhouette) : %s

            Silhouette   DBI     CH Index
K-Means     %.4f        %.4f   %.2f
Hierarchical %.4f       %.4f   %.2f
",
k_opt, best,
k_avg,  k_dbi,  k_ch,
hc_avg, hc_dbi, hc_ch))
```

Both methods identify **`r k_opt` distinct clusters** in the NYC Airbnb market. Clusters differ meaningfully in price, availability, and review activity. The best-performing method by silhouette score is highlighted in the comparison table above.
