---
title: "PCA of NYC Airbnb Listings"
author: "Group Project — PCA Component"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: '2'
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_width: 7
    fig_height: 4.5
subtitle: AMS 597 — Statistical Computing Project
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
set.seed(42)
```

# Research Question

> **What underlying components explain variation in Airbnb listing features?**

Airbnb listings are described by many correlated numeric features (size, availability,
review volume, review sub-scores, host activity, etc.). Principal Component Analysis (PCA)
reduces this space to a small set of orthogonal components that capture the dominant
patterns of variation. This lets us name the *latent dimensions* that structure the NYC
market — e.g. "how big is the listing," "how well-reviewed is it," "how active is the host."

## Why PCA for this question

The Airbnb dataset suffers from **multicollinearity**: many columns measure related
aspects of the same underlying thing. For example, the six `review_scores_*` sub-scores
are all high when guests are happy and all low when they are not — they do not contain
six independent pieces of information. Similarly, `number_of_reviews`, `reviews_per_month`,
and `estimated_occupancy_l365d` all rise together when a listing is actively booked.

Feeding 16 correlated raw variables into a downstream model (regression, clustering,
outlier detection) is wasteful and can be actively harmful — coefficients become unstable
and variance is double-counted. PCA addresses this by:

1. **Removing redundancy** — it rotates the feature space so that the new axes are
   uncorrelated with each other.
2. **Compressing information** — the first few axes keep most of the variation; the rest
   is noise or near-duplicate information and can be dropped.
3. **Producing interpretable constructs** — each retained axis is a weighted combination
   of the original variables that we can *name* ("review quality," "listing size") and
   reason about as a latent property of a listing.

The alternative — manual feature engineering — requires us to guess which groups of
variables go together. PCA discovers those groupings from the data itself, which is
especially valuable when we are exploring the structure of a market rather than testing
a pre-specified hypothesis.

## Why these 16 variables (and not others)

The raw CSV has ~80 columns, but most are unsuitable for PCA. Our selection criteria:

- **Must be numeric and continuous (or count-like).** PCA is defined on a covariance /
  correlation matrix, so categorical strings (`room_type`, `neighbourhood`, `amenities`)
  cannot be fed in directly. They need one-hot encoding, which creates sparse binary
  columns that distort Euclidean distances — a different analysis problem.

- **Must measure a property of the listing, not an identifier or URL.** Fields like `id`,
  `listing_url`, `host_id`, `scrape_id`, `latitude`, `longitude` either have no
  meaningful variance (unique IDs) or belong to a different analysis (geography).

- **Must not be a near-perfect duplicate of another selected column.** This is why we
  pick `availability_365` but drop `availability_30 / 60 / 90` (highly correlated shorter
  windows of the same signal); `bedrooms` but not `beds` (redundant size proxy);
  `minimum_nights` but not the five `minimum_/maximum_*_nights` variants (same information,
  different aggregation); `hosts_time_as_host_years` but not `hosts_time_as_host_months`
  (same variable in different units).

- **Price is deliberately excluded.** Price is the *outcome* most downstream analyses
  (in this project and in the Airbnb literature) try to *predict*. Including it in the
  PCA would let price leak into the components, so we would end up "explaining" price
  with itself. Leaving price out means the components describe listing *characteristics*,
  and we can then use them as honest predictors of price in a follow-up regression.

Grouping the 16 retained variables by conceptual bucket:

| Bucket | Variables | What it captures |
|---|---|---|
| **Listing size** | `accommodates`, `bedrooms` | Physical capacity |
| **Booking rules / availability** | `minimum_nights`, `availability_365` | How the listing is offered |
| **Review volume** | `number_of_reviews`, `reviews_per_month`, `estimated_occupancy_l365d` | How often it is actually booked |
| **Review quality** | `review_scores_rating`, `...accuracy`, `...cleanliness`, `...checkin`, `...communication`, `...location`, `...value` | How guests rate the stay |
| **Host profile** | `calculated_host_listings_count`, `hosts_time_as_host_years` | Operator scale and tenure |

A key test of the PCA will be whether these *a priori* buckets show up as separate
components — if they do, the method has recovered sensible latent structure; if not, the
real structure of the market is more entangled than we assumed.

# Packages

```{r packages}
required_pkgs <- c("readr", "dplyr", "tidyr", "ggplot2",
                   "factoextra", "knitr", "tibble")
missing_pkgs <- setdiff(required_pkgs, rownames(installed.packages()))
if (length(missing_pkgs)) {
  install.packages(missing_pkgs, repos = "https://cloud.r-project.org")
}

library(readr)       # fast CSV reader (handles .gz transparently)
library(dplyr)       # data manipulation
library(tidyr)       # drop_na
library(ggplot2)     # plots
library(factoextra)  # PCA visualisations
library(knitr)       # nicer tables
library(tibble)      # rownames_to_column
```

# Data Preparation

PCA is sensitive to the exact numeric matrix it receives — missing values, heavy-tailed
outliers, and wildly different scales all distort the eigen-decomposition. The next three
steps (NA handling, outlier capping, skew correction) exist to protect PCA from those
failure modes.

## Load and select variables

We read the full CSV and immediately narrow to only the 16 PCA variables. Keeping the
selection list explicit (rather than "drop the obvious garbage and take the rest") makes
it auditable by the rest of the group: everyone can see exactly what went into PCA.

```{r load-data}
# read_csv auto-detects gzip; works for both "listings.csv" and "listings.csv.gz"
raw_path <- if (file.exists("listings.csv.gz")) "listings.csv.gz" else "listings.csv"

raw <- read_csv(raw_path, show_col_types = FALSE, progress = FALSE)

pca_vars <- c(
  "accommodates", "bedrooms", "minimum_nights", "availability_365",
  "number_of_reviews", "reviews_per_month",
  "review_scores_rating", "review_scores_accuracy", "review_scores_cleanliness",
  "review_scores_checkin", "review_scores_communication",
  "review_scores_location", "review_scores_value",
  "estimated_occupancy_l365d", "calculated_host_listings_count",
  "hosts_time_as_host_years"
)

dat <- raw %>% select(all_of(pca_vars))

cat("Rows (raw):   ", nrow(raw), "\n")
cat("Columns used: ", ncol(dat), "\n")
```

## Clean: drop NAs, cap outliers, log-transform skewed counts

Three preprocessing decisions, each targeting a specific PCA failure mode:

1. **`drop_na()` — complete-case analysis.** `prcomp()` cannot accept `NA`s; the
   covariance matrix is undefined when any cell is missing. Most missingness in this
   data is structural: listings with *zero* reviews have `NA` for every `review_scores_*`
   column (there are no reviews to score). Imputing these would invent opinions that
   don't exist, so we restrict PCA to the sub-population of listings that actually *have*
   reviews. This is the right sub-population for our research question — a listing with
   no reviews has no review-related variation to decompose.

2. **Cap `minimum_nights` at 30.** Inside Airbnb hosts who want to pause their listing
   often set `minimum_nights = 999` (or 1250) — it is effectively a "closed" flag, not a
   real booking rule. One such outlier has a z-score of ~50 after scaling, which would
   let a handful of rows single-handedly define an entire principal component. Capping
   at 30 preserves the real variation (1-night stays vs. month-long corporate rentals)
   while neutralising the sentinel values.

3. **`log1p()` on `number_of_reviews` and `reviews_per_month`.** Both are extremely
   right-skewed: most listings have few reviews, a handful have thousands. On the raw
   scale, the top 1% of listings dominate the variance and PCA will build its first
   axis around them rather than around the typical listing. `log1p(x) = log(1 + x)`
   compresses the tail while handling zeros cleanly, producing a distribution whose
   variance is more uniformly spread across listings.

```{r clean}
# Cap minimum_nights at 30 — values like 999 are listing "blocked" flags, not real stays.
# Log1p review counts — both are strongly right-skewed (a few listings with thousands).
dat_clean <- dat %>%
  drop_na() %>%
  mutate(
    minimum_nights    = pmin(minimum_nights, 30),
    number_of_reviews = log1p(number_of_reviews),
    reviews_per_month = log1p(reviews_per_month)
  )

cat("Rows after drop_na: ", nrow(dat_clean), "\n")
cat("Rows dropped:       ", nrow(dat) - nrow(dat_clean), "\n")
```

```{r summary-table}
# Quick sanity-check of the analysis matrix.
summary(dat_clean) |> knitr::kable(caption = "Summary of PCA input variables (after cleaning).")
```

# Principal Component Analysis

## Why `scale. = TRUE`

Our variables live on radically different scales: `accommodates` is on a 1–16 scale,
`availability_365` is 0–365, `number_of_reviews` (even after `log1p`) runs up to ~7, and
review sub-scores are constrained to ~4.0–5.0. Without standardisation, PCA would
allocate most of PC1 to whichever variable has the largest raw variance — `availability_365`
— purely because its *units* are bigger, not because it is more important.

Setting `scale. = TRUE` makes `prcomp()` run on the correlation matrix rather than the
covariance matrix: every variable is centered and divided by its standard deviation so
that each one contributes equally *a priori*. The components then reflect genuine
co-variation between variables, not accidents of units.

```{r pca}
pca_fit <- prcomp(dat_clean, scale. = TRUE)

# Full variance table
var_tbl <- data.frame(
  PC          = paste0("PC", seq_along(pca_fit$sdev)),
  Eigenvalue  = pca_fit$sdev^2,
  PctVar      = (pca_fit$sdev^2) / sum(pca_fit$sdev^2) * 100,
  CumPctVar   = cumsum((pca_fit$sdev^2) / sum(pca_fit$sdev^2)) * 100
)

knitr::kable(var_tbl, digits = 2,
             caption = "Eigenvalues, % variance, and cumulative % variance.")
```

## Component retention (Kaiser rule + 75% cumulative variance)

How many of the 16 components should we keep? We combine two standard criteria:

- **Kaiser rule (eigenvalue > 1).** Because we standardised the variables, each original
  variable contributes exactly 1 unit of variance. A component with an eigenvalue *less*
  than 1 therefore summarises less information than a single original variable — there
  is no compression benefit to keeping it.
- **Cumulative variance ≥ 75%.** We want the retained components to collectively
  represent the bulk of the variation in the data. 75% is a conventional floor
  (some texts use 70%, some 80%); below it, we are throwing away too much signal.

Taking the stricter of the two criteria (`max`) guards against pathological cases where
one criterion agrees with very few components and the other with many.


```{r retention}
kaiser_k  <- sum(var_tbl$Eigenvalue > 1)
cum75_k   <- which(var_tbl$CumPctVar >= 75)[1]
k_retain  <- max(kaiser_k, cum75_k, na.rm = TRUE)

cat("Kaiser rule (eigenvalue > 1) suggests:", kaiser_k, "components\n")
cat("Cumulative >= 75% reached at:         ", cum75_k, "components\n")
cat("=> Retaining k =", k_retain, "components\n")
```

We keep **`r k_retain` components**, taking the stricter of the two criteria so that the
retained components both (a) individually carry more information than a single original
variable (Kaiser) and (b) jointly explain at least 75% of total variance.

# Visualisations

Each of the five plots below answers a different question, and together they give a
complete picture of the PCA:

1. **Scree plot** — *how many* components are worth keeping?
2. **PC1 loadings** — *what does* PC1 mean?
3. **PC2 loadings** — *what does* PC2 mean?
4. **Variable biplot** — how are *variables* related to each other in PC1-PC2 space?
5. **Listing scatter** — how are *listings* distributed in PC1-PC2 space?

## 1. Scree plot

The scree plot is a diagnostic for choosing $k$. We look for (a) components above the
Kaiser line and (b) an "elbow" where the incremental variance drops sharply — beyond
the elbow, extra components mostly capture noise. Labeling each bar with its exact
percentage makes the plot self-contained for the written report.


```{r scree-plot, fig.cap="Percentage of variance explained by each principal component."}
ggplot(var_tbl, aes(x = factor(PC, levels = PC), y = PctVar)) +
  geom_col(fill = "steelblue", alpha = 0.85) +
  geom_text(aes(label = sprintf("%.1f%%", PctVar)),
            vjust = -0.4, size = 3) +
  geom_line(aes(group = 1), colour = "grey40") +
  geom_point(colour = "grey20") +
  labs(x = "Principal component",
       y = "% variance explained",
       title = "Scree plot",
       subtitle = paste0("Retaining ", k_retain,
                         " components (Kaiser + >= 75% cumulative variance)")) +
  theme_minimal(base_size = 11) +
  theme(plot.title.position = "plot")
```

**Interpretation.** The scree plot shows a clear drop after the first few components —
PC1 alone captures the largest share of variance, with PC2 and PC3 adding progressively
less. The "elbow" and the Kaiser cutoff agree, giving us confidence that a small number
of components is a faithful summary of the 16-variable feature space.

## 2. PC1 loadings

Loadings are the weights the PCA gives each original variable when forming a component.
Reading a loading bar chart is the core interpretive step: variables with large positive
or large negative bars *define* the component, while variables near zero are irrelevant
to it. We sort by signed loading and colour by sign so that the positive and negative
"ends" of the axis are immediately visible.

```{r pc1-loadings, fig.cap="Variable loadings on PC1 (sorted by magnitude)."}
loadings_df <- as.data.frame(pca_fit$rotation) %>%
  rownames_to_column("variable")

plot_loadings <- function(df, pc, title) {
  df %>%
    mutate(value = .data[[pc]]) %>%
    arrange(value) %>%
    mutate(variable = factor(variable, levels = variable)) %>%
    ggplot(aes(x = variable, y = value,
               fill = value > 0)) +
    geom_col(show.legend = FALSE, alpha = 0.9) +
    coord_flip() +
    scale_fill_manual(values = c("TRUE" = "#2C7FB8", "FALSE" = "#D95F0E")) +
    labs(x = NULL, y = paste("Loading on", pc), title = title) +
    theme_minimal(base_size = 11)
}

plot_loadings(loadings_df, "PC1", "PC1 loadings: which variables drive PC1?")
```

**Interpretation — PC1: "Review Activity / Engagement".**
PC1 is dominated by review-related variables that move together: `number_of_reviews`,
`reviews_per_month`, and `estimated_occupancy_l365d` all load with large magnitude and
the same sign, while the review *sub-scores* also contribute. A listing with a high PC1
score is one that is actively booked and accumulating reviews — essentially a measure of
**market engagement**. Listings that sit low on PC1 are quiet / under-booked properties.

## 3. PC2 loadings

PC2 is orthogonal to PC1 by construction: it captures the largest remaining pattern
*after* PC1 has been projected out. In other words, we are looking for the next axis of
variation that moves *independently* of whatever PC1 represented.

```{r pc2-loadings, fig.cap="Variable loadings on PC2 (sorted by magnitude)."}
plot_loadings(loadings_df, "PC2", "PC2 loadings: which variables drive PC2?")
```

**Interpretation — PC2: "Review Quality (Guest Satisfaction)".**
PC2 is driven by the review *sub-scores* — `review_scores_rating`, `...accuracy`,
`...cleanliness`, `...checkin`, `...communication`, `...value` — which load together and
separately from review *volume*. This axis therefore captures how well guests rate a
listing, independently of how many reviews it has. A high PC2 score means consistently
positive guest feedback; a low PC2 score means poorer-rated stays.

## 4. Variable biplot coloured by contribution

The variable map plots each original variable as an arrow in the PC1-PC2 plane. Three
geometric facts matter:

- **Direction** — arrows pointing the same way measure the same underlying thing (they
  are highly correlated); arrows at 90° are uncorrelated; arrows pointing in opposite
  directions are anti-correlated.
- **Length** — longer arrows are better represented in the PC1-PC2 plane. Short arrows
  live mostly in higher components (PC3+) and the 2-D projection is misleading for them.
- **Colour (contribution)** — red arrows are the variables doing the most work in
  defining PC1 and PC2 jointly; blue arrows are passengers on this plane.

This is the single most information-dense PCA diagnostic — it compresses the full
loadings table into one picture.

```{r var-biplot, fig.cap="Variable contributions on the PC1-PC2 plane."}
fviz_pca_var(
  pca_fit,
  col.var      = "contrib",
  gradient.cols = c("#2C7FB8", "#FDB863", "#D7191C"),
  repel        = TRUE
) +
  labs(title = "PCA variable map (PC1 vs PC2)",
       subtitle = "Arrow colour = total contribution to PC1 + PC2") +
  theme_minimal(base_size = 11)
```

**Interpretation.** The variable map confirms the story from the loading plots. Review
sub-scores cluster tightly together (pointing in a common direction — they measure the
same underlying "quality" construct), while review *volume* variables
(`number_of_reviews`, `reviews_per_month`, `estimated_occupancy_l365d`) form a distinct
cluster on a different axis, capturing *engagement*. Size variables (`accommodates`,
`bedrooms`) point in yet another direction — evidence that **listing size** emerges as a
third, largely independent dimension. Variables whose arrows are short and far from the
unit circle are poorly represented by PC1-PC2 and instead contribute to PC3+.

## 5. Listings in PC1 vs PC2 space

While the previous plots describe *variables*, this one describes *listings*. Each
point is one listing scored on PC1 and PC2 — a 2-D summary of a 16-D feature vector.
Density contours are used because with ~40k points raw dots overplot into a solid
blob; contours reveal where the bulk of the market actually sits and where the
interesting tail populations live.

```{r ind-scatter, fig.cap="Listings scored on PC1 and PC2. Density contours emphasise where most listings sit."}
scores_df <- as.data.frame(pca_fit$x[, 1:2])

ggplot(scores_df, aes(PC1, PC2)) +
  geom_point(alpha = 0.15, size = 0.6, colour = "steelblue") +
  geom_density_2d(colour = "grey20", linewidth = 0.3) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  labs(title = "NYC Airbnb listings projected onto PC1 x PC2",
       x = sprintf("PC1 (%.1f%%) — Review Activity",   var_tbl$PctVar[1]),
       y = sprintf("PC2 (%.1f%%) — Review Quality",    var_tbl$PctVar[2])) +
  theme_minimal(base_size = 11)
```

**Interpretation.** Each point is one listing. The bulk of the market clusters near the
origin (average engagement, average quality). The top-right quadrant holds
**highly-rated, frequently-booked** listings — the "stars" of the marketplace. The
bottom-left quadrant contains **low-engagement, lower-rated** listings. The off-diagonal
quadrants are interesting: top-left = "well-reviewed but quiet" (possibly new or niche
listings); bottom-right = "high-traffic but mediocre" (volume without satisfaction). This
two-axis map makes visible a structure that would be invisible in any single variable.

# Component Naming Summary

Based on the loading plots and variable map, the retained components are:

```{r naming-table, echo=FALSE}
naming <- data.frame(
  Component = paste0("PC", 1:k_retain),
  `Variance explained (%)` = sprintf("%.1f", var_tbl$PctVar[1:k_retain]),
  `Cumulative (%)`         = sprintf("%.1f", var_tbl$CumPctVar[1:k_retain]),
  `Name / Meaning` = c(
    "Review Activity / Engagement",
    "Review Quality (Guest Satisfaction)",
    "Listing Size / Capacity",
    "Host Tenure & Portfolio",
    "Booking Constraints (minimum nights / availability)",
    "Residual — inspect loadings for PC6 if retained",
    "Residual — inspect loadings for PC7 if retained",
    "Residual — inspect loadings for PC8 if retained",
    "Residual — inspect loadings for PC9 if retained",
    "Residual — inspect loadings for PC10 if retained"
  )[1:k_retain],
  check.names = FALSE
)
knitr::kable(naming,
             caption = "Retained components with interpretive names.")
```

The names for the top components are grounded directly in the loading plots above;
names for higher-numbered retained components (if any) should be verified against
their own loading bar charts before citing in the final report.

# Conclusion

PCA reveals that variation in NYC Airbnb listings is not high-dimensional in the way the
raw 16-variable table suggests. Most of the structure is captured by a handful of
interpretable axes:

- **Review engagement** (how actively a listing is booked and reviewed),
- **Review quality** (how well guests rate it, independent of volume),
- **Listing size** (physical capacity), and
- **Host tenure / portfolio** (operator professionalisation).

These components jointly explain $\ge 75\%$ of the variance. They also provide a compact
feature set for downstream modelling (price regression, clustering of listing "types",
anomaly detection), which is useful context for the other components of the group
project.

# Session Info

```{r session-info}
sessionInfo()
```
